\chapter*{Abstract}
This dissertation addresses the complex challenges and solutions associated with the implementation and testing of consensus algorithms in blockchain networks. The research is anchored on the Tezos blockchain platform, selected for its adaptability and versatility. The primary aim is to advance the field of blockchain technology by developing tools and methods that facilitate the easy swapping, testing, and comparative analysis of different consensus algorithms in a live network environment.

The document is organized into several pivotal sections. It begins with an introduction that outlines the problem and solution, emphasizing the existing gaps in the development and testing of consensus algorithms, particularly the absence of tools for live testing and the challenges in swapping algorithms seamlessly. This is followed by a comprehensive review of the current state-of-the-art in consensus algorithms, offering a detailed analysis of their efficiency, limitations, and suitability for various applications.

A preliminary case study is then presented, focusing on the implementation of a Proof of Work (PoW) consensus algorithm on the Tezos platform. This serves as a proof of concept and provides valuable insights into the inner workings of consensus algorithms and their integration with existing blockchain nodes.

The core contribution of this dissertation is the development of two tools: one aimed at easing the development of consensus algorithms and another designed for their live testing. These tools are intended to serve both the academic and industrial communities interested in blockchain technology.

Finally, the dissertation presents a real-world scenario that employs both tools. A Proof of Authority (PoA) algorithm is implemented using the development tool and then tested using the live testing tool. The results are compared with the previously implemented PoW algorithm, providing empirical evidence of the efficiency and limitations of both algorithms.

In summary, this dissertation offers a robust framework for the development and testing of consensus algorithms, substantiated by empirical evidence from the Tezos platform. It holds significant implications for both academic researchers and industry stakeholders, laying a strong foundation for future work in this rapidly evolving field


\section*{Keywords}
Blockchain, Consensus Algorithms, Tezos Platform, Proof of Work (PoW), Proof of Authority (PoA), Live Testing, Protocol Development
